---
title: Reference
slug: reference
output:
  blogdown::html_page:
    template: ../pandoc/toc-title_html.template
    toc: true
---

```{r include=FALSE, eval=F}
library(tidyverse)
library(emo)
library(here)

source(here("lib", "graphics.R"))
```

# Installing R and R Studio

1. [Install R](http://www.r-project.org/) from CRAN^[The Comprehensive R Archive Network] by clicking "Download R" (or the CRAN link under Downloads on the left menu). This will take you to a mirrors page, where you can select a location in the U.S. and download a copy of R
2. [Install R Studio (Desktop Version)](https://www.rstudio.com/products/rstudio/download/), choose the "Free" option

# R Packages

Packages come from multiple sources. 

The polished, publicly released versions are found on [CRAN](https://cran.r-project.org/). When installing a package available on CRAN, it is sufficient simply to tell R the following:^[Note the plural `s` on `packages`, and the quotes around the "package name"]

```{r, eval=F}
install.packages("packagename") 
```

Other packages, which may be in various developmental states (including perfectly functional!) are often hosted on [GitHub](http://github.com) before they make their way onto CRAN. Simply telling R `install.packages("packagename")` will fail to find it (as R only looks in CRAN for packages), so you must use another package called `devtools`^[Which you will need to install first if you (probably) don't already have it!] to install packages directly from Github:^[Note the `::` allows you to use the function `install_github()` from the `devtools` package without having to first load the devtools package with `library(devtools)`.]

```{r, eval=F}
devtools::install_github("username/packagename") 
```

For example, to install Hadley Wickham's package `r4ds` from its Github page [https://github.com/hadley/r4ds](https://github.com/hadley/r4ds), we would type:

```{r, eval=F}
devtools::install_github("hadley/r4ds")
```

# Getting Help for R

For specific functions or commands, you can simply type:

```{r, eval=F}
?functionname()

# example
?mean()
```

This will display a help page specific to that function in the Viewer pane. R functions and packages are extremely well-documented; help pages normally include a short description of the function, arguments and options (as well as their default values), and several examples or vignettes to demonstrate usage of the function. 

Additionally, you can turn to the community. 

# R Markdown 

## Math and $\LaTeX{}$

Math in R Markdown uses the $\LaTeX{}$ language to typeset beautiful formulas and equations. To include mathematical symbols or expressions inline^[That means, within the text, and not in a separate line, location, or environment within the document.], insert it `$`between two dollar signs`$`. Within a sentence, `$2^2+\frac{\pi}{\pi}=5$` becomes $2^2+\frac{\pi}{\pi}=5$. 

If you prefer it to be centered in its own line, put it on its own line, `$$`between two dollar signs`$$`. 

`$``$``2^2+\frac{\pi}{\pi}=5``$``$`` becomes:

$$2^2+\frac{\pi}{\pi}=5$$ 

There is also a shortcut with the R package `equatiomatic`. 

## Producing documents

### Producing a PDF Output

PDFs are archaic document formats [with many flaws](https://www.urban.org/urban-wire/your-data-deserve-better-pdf), yet the format remains the industry standard in most cases. PDFs are made with $\LaTeX{}$, a typography language dating to the 1980s. $\LaTeX{}$ [has its own problems](https://www.authorea.com/users/5713/articles/19359-latex-is-dead-long-live-latex-typesetting-in-the-digital-age/_show_article) 

One alternative is [tinytex](https://yihui.name/tinytex/)

```{r, eval=F}
install.packages("tinytex")
tinytex::install_tinytex()  # install TinyTeX
```

# Resources

## Cheat Sheets



# R Packages

The following is a compendium of all R packages used, and their main uses

- `tidyverse`
- `DT`
- `knitr`
- `haven`
- `readxl`
- `dplyr`
- `ggplot2`
- `car`
- `sandwich`
- `lmtest`
- `equatiomatic`
- `gganimate`
- `stargazer`
- `broom`
- `modelr`
- `infer`
- `aer`

# Finding Data

- A [near-comprehensive list](https://vincentarelbundock.github.io/Rdatasets/datasets.html) of all existing data sets built-in to R or R packages^[Note: You should use these more for playing around with in R to boost your data wrangling skills. These should not be used for your projects in most circumstances.]

# Citations and bibliography

[^bibtex] You can [download a BibTeX file](/bib/references.bib) of all the non-web-based readings in the course.

[^bibtex]: {-} You can open the file in [BibDesk on macOS](http://bibdesk.sourceforge.net/), [JabRef on Windows](http://www.jabref.org/), or [Zotero](https://www.zotero.org/) or [Mendeley](https://www.mendeley.com/) online.